### 2017-02-03 Thesis/latex/paperA.tex
TBD: paperB, paperC, bibliography

In its simplest form, the water-wave problem concerns the
flow of an incompressible inviscid fluid with a free surface over
a horizontal impenetrable bed. In this situation, the fluid flow
is described by the Euler equations with appropriate boundary
conditions, and the dynamics of the free surface are of particular
interest in the solution of this problem.
There are a number of model equations which allow the
approximate description of the evolution of the free surface
without having to provide a complete solution of the fluid flow
below the surface. In the present contribution, interest is focused

on the derivation and evaluation of a nonlocal water-wave model
known as the Whitham equation. The equation is written as

\begin{equation}
\eta_t + \frac{3}{2} \frac{c_0}{h_0} \eta \eta_x + K_{h_0} \ast \eta_x = 0, 
\label{whitham-equation}
\end{equation}
where the convolution kernel $K_{h_0}$ is given in terms of the Fourier
transform by

\begin{equation}
\mathcal{F}K_{h_0} (\xi) =  \sqrt{\frac{g \tanh(h_0\xi)}{\xi}}.
\label{whitham-kernel}
\end{equation}
$g$ is the gravitational acceleration, $h_0$ is the undisturbed depth of
the fluid, and $c_0 = \sqrt{gh_0}$ is the corresponding long-wave speed.
The convolution can be thought of as a Fourier multiplier operator,
and \eqref{whitham-kernel} represents the Fourier symbol of the operator.
The Whitham equation was proposed by Whitham \cite{Whitham1967} as an
alternative to the well known Korteweg–de Vries (KdV) equation

\begin{equation}
\eta_t + c_0 \eta_x + \frac{3}{2} \frac{c_0}{h_0} \eta \eta_x + \frac{1}{6} c_0 h_0^2 \eta_{xxx} = 0.
\label{kdv-equation}
\end{equation}

The validity of the KdV equation as a model for surface water waves
can be described as follows. Suppose a wave field with a prominent
amplitude $a$ and characteristic wavelength $l$ is to be studied. The
KdV equation is known to produce a good approximation of the
evolution of the waves if the amplitude of the waves is small and
the wavelength is large when compared to the undisturbed depth,
and if in addition, the two non-dimensional quantities $a / h_0$ and
$h_0^2 / l^2$ are of similar size. The latter requirement can be written in
terms of the Stokes number as

\begin{equation}
\mathcal{S} = \frac{a l^2}{h_0^3} \sim 1. 
\label{stockes-number}
\end{equation}
While the KdV equation is a good model for surface waves if $\mathcal{S} \sim 1$, 
one notorious problem with the KdV equation is that it does
not model accurately the dynamics of shorter waves. Recognizing
this shortcoming of the KdV equation, Whitham proposed to use
the same nonlinearity as the KdV equation, but couple it with a
linear term which mimics the linear dispersion relation of the full
water-wave problem. Thus, at least in theory, the Whitham
equation can be expected to yield a description of the dynamics
of shorter waves which is closer to the solutions of the more
fundamental Euler equations which govern the flow.

The Whitham equation has been studied from a number of
vantage points during recent years. In particular, the existence of
traveling and solitary waves has been established in \citep{Ehrnstrom2012, Ehrnstrom2009}.
Well
posedness of a similar equation was investigated in \citetext{\citealp[]{Lannes2013}-\citealp[]{Klein2015}}, and a
model with variable depth has been studied numerically in \cite{Minzoni2013}.
Moreover, it has been shown in \citep{Hur2015, Sanford2014} that periodic solutions of
the Whitham equation feature modulational instability for short
enough waves in a similar way as small-amplitude periodic wave
solutions of the water-wave problem. However, even though the
equation is routinely mentioned in texts on nonlinear waves \citep{Drazin89, Whitham74}, it appears that the performance of the Whitham equation in
the description of surface water waves has not been investigated
so far.
The purpose of the present article is to give an asymptotic
derivation of the Whitham equation as a model for surface water
waves, and to confirm Whitham’s expectation that the equation is
a fair model for the description of time-dependent surface water
waves. For the purpose of the derivation, we introduce an expo-
nential scaling regime in which the Whitham equation can be de-
rived asymptotically from an approximate Hamiltonian principle
for surface water waves. In order to motivate the use of this scaling,
note that the KdV equation has the property that wide classes of
initial data decompose into a number of solitary waves and small-
amplitude dispersive residue \cite{Ablowitz81}. For the KdV equations, solitary-
wave solutions are known in closed form, and are given by
\begin{equation}
\eta = \frac{a}{h_0}\mbox{sech}^2\left(\sqrt{\frac{3a}{4h_0^3}(x - ct)}\right)
\label{kdv-soliton}
\end{equation}
for a certain wave celerity $c$. These waves clearly comply with
the amplitude–wavelength relation $a / h_0 \sim h_0^2 / l^2$ which was
mentioned above. It appears that the Whitham equation – as indeed do many other nonlinear dispersive equations – also has
the property that broad classes of initial data rapidly decompose
into ordered trains of solitary waves (see Fig. 1). Quantifying the
amplitude–wavelength relation for these solitary waves yields an
asymptotic regime which is expected to be relevant to the validity
of the Whitham equation as a water wave model.

As the curve fit in the right panel of Fig. 1 shows, the relationship
between wavelength and amplitude of the Whitham solitary
waves can be approximately described by the relation $a/h_0 \sim e^{\kappa(l/h_0)^{\nu}}$ for certain values of $\kappa$ and $\nu$. Since the Whitham solitary
waves are not known in exact form, the values of $\kappa$ and $\nu$ have to be
found numerically. Then one may define a Whitham scaling regime
\begin{equation}
\mathcal{W}(\kappa, \nu) = \frac{a}{h_0}e^{\kappa(l/h_0)^{\nu}} \sim 1,
\label{whitham-regime}
\end{equation}
and it will be shown in Sections 2 and 3 that this scaling can be used
advantageously in the derivation of the Whitham equation. The
derivation proceeds by examining the Hamiltonian formulation of
the water-wave problem due to Zakharov, Craig and Sulem \citep{Craig1993, Zakharov1968},
and by restricting to wave motion which is predominantly in the
direction of increasing values of x. The approach is similar to the
method of \cite{Craig1994}, but relies on the new relation (5).
First, in Section 2, a Whitham system is derived which allows for
two-way propagation of waves. The Whitham equation is found in
Section 3. Finally, in Section 4, a comparison of modeling properties
of the KdV and Whitham equations is given. The comparison also
includes the regularized long-wave equation
\begin{equation}
\eta_t + c_0 \eta_x + \frac{3}{2} \frac{c_0}{h_0} \eta \eta_x - \frac{1}{6} h_0^2 \eta_{xxt} = 0,
\label{regularized-long-wave}
\end{equation}
which was put forward in \cite{Peregrine1966} and studied in depth in \cite{Benjamin1972}, and
which is also known as the BBM or PBBM equation. The linearized
dispersion relation of this equation is not an exact match to the
dispersion relation of the full water-wave problem, but it is much
closer than the KdV equation, and it might also be expected that
this equation may be able to model shorter waves more success-
fully than the KdV equation. In order to obtain an even better match
of the linear dispersion relation, one may make use of Pad\'e  expansions. In the context of simplified evolution equations, this
approach was pioneered in \cite{Witting1984}. For uni-directional models, this
approach was advocated in \cite{Fetecau2005}, and in particular, the equation
based on the Pad\'e (2,2) approximation was studied in depth. In dimensional variables, this model takes the form
\begin{equation}
\eta_t + c_0 \eta_x + \frac{3}{2} \frac{c_0}{h_0} \eta \eta_x
- \frac{3}{20} c_0 h_0^2 \eta_{xxx} - \frac{19}{60} h_0^2 \eta_{xxt} = 0.
\label{pade-2-2}
\end{equation}
The dispersion relations for the KdV, BBM and Pad\'e (2,2) models are respectively
\begin{align*}
c(k) & = c_0 - \frac{1}{6} c_0 h_0^2 k^2 \qquad \mbox{(KdV)}, \\
c(k) & = c_0 \frac{1}{1 + \frac{1}{6} h_0^2 k^2} \qquad ~\mbox{(BBM)}, \\
c(k) & = c_0 \frac{1 + \frac{3}{20} h_0^2 k^2}{1 + \frac{19}{60} h_0^2 k^2} \qquad \mbox{(Pad\'e (2,2))}. \\
\label{dispersive-relations}
\end{align*}

These approximate dispersion relations are compared to the full
dispersion relation in Fig. 2. It appears clearly that the Pad\'e (2,2)
approximation remains much closer to the full dispersion relation than the dispersion relations based on either the linear KdV
or linear BBM equations. As will be seen in Section 4, solutions of
both the Whitham and Pad\'e (2,2) equations give closer approximations to solutions of the full Euler equations than either the KdV or
BBM equations in most cases investigated. However, the Whitham
equation still keeps a slight edge over the Pad\'e model.

@@

It will be convenient for the present purpose to formulate the
Hamiltonian in terms of the dependent variable $u = \Phi_x$. To this
end, we define the operator $\mathcal{K} (\eta)$ by
%
\begin{equation*}
	G(\eta) = D \mathcal{K}(\eta) D
	.
\end{equation*}
%
As was the case with $G (\eta)$, the operator $\mathcal{K}(\eta)$ can be expanded in
a Taylor series around zero as
%
\[
	\mathcal{K}(\eta) = \sum_{j=0}^\infty \mathcal{K}_j(\eta)
	, \qquad
	\mathcal{K}_j(\eta) = D^{-1}G_j(\eta)D^{-1}
	.
\]
%
In particular, note that $\mathcal{K}_0 = \frac{\tanh D}{D}$. In non-dimensional variables, we write the operator with the integral kernel $K_{h_0}$ as $K = \sqrt{\frac{\tanh D}{D}}$, so that we have $\mathcal{K}_0 = K^2$. The Hamiltonian is then expressed as 
%
\begin{equation}
	H  = \frac{1}{2} \int_{\mathbb{R}} \left[ \eta^2 + u \mathcal{K}(\eta) u \right] \, dx.
\label{scaled-Hamiltonian}
\end{equation}
%
The water-wave problem can then be written as a Hamiltonian
system using the variational derivatives of $H$ and posing the
Hamiltonian equations
%
\begin{equation}
	\eta_t = - \partial_x \frac{\delta H}{\delta u}, \qquad u_t = -\partial_x \frac{\delta H}{\delta \eta}.
\end{equation}
%
This system is not in canonical form as the associated structure
map $J_{\eta, u}$ is symmetric:
%
\begin{equation}
	J_{\eta, u} =
	\begin{pmatrix}
		0 & -\partial_x
		\\
		-\partial_x & 0
	\end{pmatrix}.
\end{equation}
%
We now proceed to derive a system of equations which is similar to
the Whitham equation \eqref{whitham-equation}, but allows bi-directional wave propagation. This system will be a stepping stone on the way to a derivation
of \eqref{whitham-equation}, but may also be of independent interest. Consider a wave-
field having a characteristic wavelength $l$ and a characteristic amplitude $a$. Taking into account the nondimensionalization, the two
scalar parameters $\lambda = l / h_0$ and $ \alpha = a / h_0$ appear. In order to introduce the long-wave and small amplitude approximation into the
non-dimensional problem, we use the scaling $\tilde{x}= \frac{1}{\lambda} x$, and $ \eta = \alpha \tilde{\eta}$.
This induces the transformation $\tilde{D} = \lambda D = -\lambda i \partial_x $. If the energy
is nondimensionalized in accord with the nondimensionalization
mentioned earlier, then the natural scaling for the Hamiltonian is
$ \tilde{H}= \alpha^2 H$. In addition, the unknown $u$ is scaled as $u = \alpha \tilde{u}$. The
scaled Hamiltonian \eqref{scaled-Hamiltonian} is then written as

>>>>

In its simplest form, the water-wave problem concerns the
flow of an incompressible inviscid fluid with a free surface over
a horizontal impenetrable bed. In this situation, the fluid flow
is described by the Euler equations with appropriate boundary
conditions, and the dynamics of the free surface are of particular
interest in the solution of this problem.

There are a number of model equations which allow the
approximate description of the evolution of the free surface
without having to provide a complete solution of the fluid flow
below the surface. In the present contribution, interest is focused
on the derivation and evaluation of a nonlocal water-wave model
known as the Whitham equation. The equation is written as
%
\begin{equation}
	\eta_t + \frac{3}{2} \frac{c_0}{h_0} \eta \eta_x + K_{h_0} \ast \eta_x = 0, 
	\label{whitham-equation}
\end{equation}
%
where the convolution kernel $K_{h_0}$ is given in terms of the Fourier
transform by
%
\begin{equation}
	\mathcal{F}K_{h_0} (\xi) =  \sqrt{\frac{g \tanh(h_0\xi)}{\xi}}.
	\label{whitham-kernel}
\end{equation}
%
$g$ is the gravitational acceleration, $h_0$ is the undisturbed depth of
the fluid, and $c_0 = \sqrt{gh_0}$ is the corresponding long-wave speed.
The convolution can be thought of as a Fourier multiplier operator,
and \eqref{whitham-kernel} represents the Fourier symbol of the operator.

The Whitham equation was proposed by Whitham \cite{Whitham1967} as an
alternative to the well known Korteweg–de Vries (KdV) equation
%
\begin{equation}
	\eta_t + c_0 \eta_x + \frac{3}{2} \frac{c_0}{h_0} \eta \eta_x + \frac{1}{6} c_0 h_0^2 		\eta_{xxx} = 0.
	\label{kdv-equation}
\end{equation}
%
The validity of the KdV equation as a model for surface water waves
can be described as follows. Suppose a wave field with a prominent
amplitude $a$ and characteristic wavelength $l$ is to be studied. The
KdV equation is known to produce a good approximation of the
evolution of the waves if the amplitude of the waves is small and
the wavelength is large when compared to the undisturbed depth,
and if in addition, the two non-dimensional quantities $a / h_0$ and
$h_0^2 / l^2$ are of similar size. The latter requirement can be written in
terms of the Stokes number as
%
\begin{equation}
\mathcal{S} = \frac{a l^2}{h_0^3} \sim 1. 
\label{stockes-number}
\end{equation}
%
While the KdV equation is a good model for surface waves if $\mathcal{S} \sim 1$, 
one notorious problem with the KdV equation is that it does
not model accurately the dynamics of shorter waves. Recognizing
this shortcoming of the KdV equation, Whitham proposed to use
the same nonlinearity as the KdV equation, but couple it with a
linear term which mimics the linear dispersion relation of the full
water-wave problem. Thus, at least in theory, the Whitham
equation can be expected to yield a description of the dynamics
of shorter waves which is closer to the solutions of the more
fundamental Euler equations which govern the flow.

The Whitham equation has been studied from a number of
vantage points during recent years. In particular, the existence of
traveling and solitary waves has been established in \citep{Ehrnstrom2012, Ehrnstrom2009}.
Well posedness of a similar equation was investigated in \citetext{\citealp[]{Lannes2013}-\citealp[]{Klein2015}}, and a
model with variable depth has been studied numerically in \cite{Minzoni2013}.
Moreover, it has been shown in \citep{Hur2015, Sanford2014} that periodic solutions of
the Whitham equation feature modulational instability for short
enough waves in a similar way as small-amplitude periodic wave
solutions of the water-wave problem. However, even though the
equation is routinely mentioned in texts on nonlinear waves \citep{Drazin89, Whitham74}, it appears that the performance of the Whitham equation in
the description of surface water waves has not been investigated
so far.

The purpose of the present article is to give an asymptotic
derivation of the Whitham equation as a model for surface water
waves, and to confirm Whitham’s expectation that the equation is
a fair model for the description of time-dependent surface water
waves. For the purpose of the derivation, we introduce an exponential scaling regime in which the Whitham equation can be derived asymptotically from an approximate Hamiltonian principle
for surface water waves. In order to motivate the use of this scaling,
note that the KdV equation has the property that wide classes of
initial data decompose into a number of solitary waves and small-
amplitude dispersive residue \cite{Ablowitz81}. For the KdV equations, solitary-
wave solutions are known in closed form, and are given by
%
\begin{equation}
	\eta = \frac{a}{h_0}\mbox{sech}^2\left(\sqrt{\frac{3a}{4h_0^3}(x - ct)}\right)
	\label{kdv-soliton}
\end{equation}
%
for a certain wave celerity $c$. These waves clearly comply with
the amplitude–wavelength relation $a / h_0 \sim h_0^2 / l^2$ which was
mentioned above. It appears that the Whitham equation – as indeed do many other nonlinear dispersive equations – also has
the property that broad classes of initial data rapidly decompose
into ordered trains of solitary waves (see Fig. 1). Quantifying the
amplitude–wavelength relation for these solitary waves yields an
asymptotic regime which is expected to be relevant to the validity
of the Whitham equation as a water wave model.

As the curve fit in the right panel of Fig. 1 shows, the relationship
between wavelength and amplitude of the Whitham solitary
waves can be approximately described by the relation $a/h_0 \sim e^{\kappa(l/h_0)^{\nu}}$ for certain values of $\kappa$ and $\nu$. Since the Whitham solitary
waves are not known in exact form, the values of $\kappa$ and $\nu$ have to be
found numerically. Then one may define a Whitham scaling regime
%
\begin{equation}
	\mathcal{W}(\kappa, \nu) = \frac{a}{h_0}e^{\kappa(l/h_0)^{\nu}} \sim 1,
	\label{whitham-regime}
\end{equation}
%
and it will be shown in Sections 2 and 3 that this scaling can be used
advantageously in the derivation of the Whitham equation. The
derivation proceeds by examining the Hamiltonian formulation of
the water-wave problem due to Zakharov, Craig and Sulem \citep{Craig1993, Zakharov1968},
and by restricting to wave motion which is predominantly in the
direction of increasing values of x. The approach is similar to the
method of \cite{Craig1994}, but relies on the new relation \eqref{whitham-regime}.

First, in Section 2, a Whitham system is derived which allows for
two-way propagation of waves. The Whitham equation is found in
Section 3. Finally, in Section 4, a comparison of modeling properties
of the KdV and Whitham equations is given. The comparison also
includes the regularized long-wave equation
%
\begin{equation}
	\eta_t + c_0 \eta_x + \frac{3}{2} \frac{c_0}{h_0} \eta \eta_x - \frac{1}{6} h_0^2 			\eta_{xxt} = 0,
	\label{regularized-long-wave}
\end{equation}
%
which was put forward in \cite{Peregrine1966} and studied in depth in \cite{Benjamin1972}, and
which is also known as the BBM or PBBM equation. The linearized
dispersion relation of this equation is not an exact match to the
dispersion relation of the full water-wave problem, but it is much
closer than the KdV equation, and it might also be expected that
this equation may be able to model shorter waves more success-
fully than the KdV equation. In order to obtain an even better match
of the linear dispersion relation, one may make use of Pad\'e  expansions. In the context of simplified evolution equations, this
approach was pioneered in \cite{Witting1984}. For uni-directional models, this
approach was advocated in \cite{Fetecau2005}, and in particular, the equation
based on the Pad\'e (2,2) approximation was studied in depth. In dimensional variables, this model takes the form
%
\begin{equation}
	\eta_t + c_0 \eta_x + \frac{3}{2} \frac{c_0}{h_0} \eta \eta_x
	- \frac{3}{20} c_0 h_0^2 \eta_{xxx} - \frac{19}{60} h_0^2 \eta_{xxt} = 0.
	\label{pade-2-2}
\end{equation}
%
The dispersion relations for the KdV, BBM and Pad\'e (2,2) models are respectively
%
\begin{align*}
	c(k) & = c_0 - \frac{1}{6} c_0 h_0^2 k^2 \qquad \mbox{(KdV)}, \\
	c(k) & = c_0 \frac{1}{1 + \frac{1}{6} h_0^2 k^2} \qquad ~\mbox{(BBM)}, \\
	c(k) & = c_0 \frac{1 + \frac{3}{20} h_0^2 k^2}{1 + \frac{19}{60} h_0^2 k^2} \qquad 			\mbox{(Pad\'e (2,2))}.
	\label{dispersive-relations}
\end{align*}
%
These approximate dispersion relations are compared to the full
dispersion relation in Fig. 2. It appears clearly that the Pad\'e (2,2)
approximation remains much closer to the full dispersion relation than the dispersion relations based on either the linear KdV
or linear BBM equations. As will be seen in Section 4, solutions of
both the Whitham and Pad\'e (2,2) equations give closer approximations to solutions of the full Euler equations than either the KdV or
BBM equations in most cases investigated. However, the Whitham
equation still keeps a slight edge over the Pad\'e model.

@@

It will be convenient for the present purpose to formulate the
Hamiltonian in terms of the dependent variable $u = \Phi_x$. To this
end, we define the operator $\mathcal{K} (\eta)$ by
%
\begin{equation*}
	G(\eta) = D \mathcal{K}(\eta) D
	.
\end{equation*}
%
As was the case with $G (\eta)$, the operator $\mathcal{K}(\eta)$ can be expanded in
a Taylor series around zero as
%
\[
	\mathcal{K}(\eta) = \sum_{j=0}^\infty \mathcal{K}_j(\eta)
	, \qquad
	\mathcal{K}_j(\eta) = D^{-1}G_j(\eta)D^{-1}
	.
\]
%
In particular, note that $\mathcal{K}_0 = \frac{\tanh D}{D}$. In non-dimensional variables, we write the operator with the integral kernel $K_{h_0}$ as $K = \sqrt{\frac{\tanh D}{D}}$, so that we have $\mathcal{K}_0 = K^2$. The Hamiltonian is then expressed as 
%
\begin{equation}
	H  = \frac{1}{2} \int_{\mathbb{R}} \left[ \eta^2 + u \mathcal{K}(\eta) u \right] \, dx.
	\label{scaled-Hamiltonian}
\end{equation}
%
The water-wave problem can then be written as a Hamiltonian
system using the variational derivatives of $H$ and posing the
Hamiltonian equations
%
\begin{equation}
	\eta_t = - \partial_x \frac{\delta H}{\delta u}, \qquad u_t = -\partial_x \frac{\delta H}{\delta \eta}.
\end{equation}
%
This system is not in canonical form as the associated structure
map $J_{\eta, u}$ is symmetric:
%
\begin{equation}
	J_{\eta, u} =
	\begin{pmatrix}
		0 & -\partial_x
		\\
		-\partial_x & 0
	\end{pmatrix}.
\end{equation}
%
We now proceed to derive a system of equations which is similar to
the Whitham equation \eqref{whitham-equation}, but allows bi-directional wave propagation. This system will be a stepping stone on the way to a derivation
of \eqref{whitham-equation}, but may also be of independent interest. Consider a wave-
field having a characteristic wavelength $l$ and a characteristic amplitude $a$. Taking into account the nondimensionalization, the two
scalar parameters $\lambda = l / h_0$ and $ \alpha = a / h_0$ appear. In order to introduce the long-wave and small amplitude approximation into the
non-dimensional problem, we use the scaling $\tilde{x}= \frac{1}{\lambda} x$, and $ \eta = \alpha \tilde{\eta}$.
This induces the transformation $\tilde{D} = \lambda D = -\lambda i \partial_x $. If the energy
is nondimensionalized in accord with the nondimensionalization
mentioned earlier, then the natural scaling for the Hamiltonian is
$ \tilde{H}= \alpha^2 H$. In addition, the unknown $u$ is scaled as $u = \alpha \tilde{u}$. The
scaled Hamiltonian \eqref{scaled-Hamiltonian} is then written as
%
\begin{multline*}
	\tilde{H} = \frac{1}{2} \int_{\R} \teta^2 \, dx
 + \frac{1}{2} \int_{\R} \tilde{u} \left[ 1 - \Sfrac{1}{3} \lambda^{-2} \tilde{D}^2 + \cdots \right] \tilde{u} \, dx  
 + \frac{\alpha}{2} \int_{\R}  \teta  \tilde{u}^2 \, dx
\\
 - \frac{\alpha}{2} \int_{\R}  \tilde{u} \left[ \lambda^{-1} \tilde{D} - \Sfrac{1}{3} \lambda^{-3}\tilde{D}^3 +  \cdots \right] \teta 
   \left[ \lambda^{-1} \tilde{D} - \Sfrac{1}{3} \lambda^{-3}\tilde{D}^3 + \cdots \right] \tilde{u} \, dx.
\end{multline*}
%
Let us now introduce the small parameter $\mu= \frac{1}{\lambda}$,
and assume for simplicity that $\alpha = e^{- \kappa / \mu^\nu}$, which corresponds to the
case where $\mathcal{W}(\kappa,\nu) = 1$. 
Then the Hamiltonian can be written in the following form:
%
\begin{multline*}
\tH = \ohf \int_{\R} \teta^2 \, dx
        + \ohf \int_{\R} \tu \left[ 1 - \Sfrac{1}{3} \mu^{2} \tD^2 + \cdots \right] \tu \, dx  
+ \frac{e^{-\kappa/\mu^\nu}}{2} \int_{\R}  \teta  \tu^2 \, dx
\\
- \frac{e^{-\kappa/\mu^\nu}}{2} \int_{\R}  \tu \left[ \mu \tD - \Sfrac{1}{3} \mu^{3}\tD^3 +  \dots\right]
                      \teta \left[ \mu \tD - \Sfrac{1}{3} \mu^{3}\tD^3 + \cdots \right] \tu \, dx.
\end{multline*}
%
Disregarding terms of order $\scrO(\mu^2 e^{-\kappa/\mu^\nu})$, but not of order
$\scrO(e^{\sca})$ yields the expansion
\begin{equation}
\tH = \ohf \int_{\R} \teta^2 \, dx 
    + \ohf \int_{\R} \tu \left[ 1 - \Sfrac{1}{3} \mu^{2} \tD^2 + \dots \right] \tu \, dx  
    + \frac{e^{\sca}}{2} \int_{\R}  \teta  \tu^2 \, dx.   
\end{equation}
%
Note that by taking $\mu$ small enough, an arbitrary number of terms of algebraic order 
in $\mu$ may be kept in the asymptotic series, 
so that the truncated version of the Hamiltonian in dimensional 
variables may be written as
%
\begin{equation}\label{Hamiltonian-eta-u-truncated}
	H   = \ohf \int_\R \left[ \eta^2 + u \mathcal{K}_0^N (\eta) u + u \eta u \right] \, dx dz.
\end{equation}
%
However, the difference between $\mathcal{K}_0$ and $\mathcal{K}_0^N$
is below the order of approximation, so that it is possible
to formally define the truncated Hamiltonian with
$\mathcal{K}_0$ instead of $\mathcal{K}_0^N$.
%
Hence, the Whitham system is obtained from the 
Hamiltonian \eqref{Hamiltonian-eta-u-truncated} as follows:
%
\begin{align}
	\label{sys1}
	\eta_t = - \partial_x \frac{\delta H}{\delta u} 
       &= -  \mathcal{K}_0 u_x - (\eta u)_x, \\
	\label{sys2}
	u_t = -\partial_x \frac{\delta H}{\delta \eta}
       &= - \eta_x - u u_x.
\end{align}
%
One may also derive a higher-order equation by keeping terms of order 
$\scrO(\mu^2 e^{\sca})$, but discarding terms of order $\scrO(\mu^4 e^{\sca})$.
In this case we find the system
%
\begin{align*}
	\eta_t & = -  \mathcal{K}_0 u_x - (\eta u)_x - (\eta u_x)_{xx}, \\
	u_t    & = - \eta_x - u u_x + u_x u_{xx}.
\end{align*}
%
% *************************************************************************************
\subsection{Derivation of evolution equations of Whitham type.}
% *************************************************************************************
%
In order to derive the Whitham equation for uni-directional wave propagation,
it is important to understand how solutions of the Whitham system \eqref{sys1}-\eqref{sys2}
can be restricted to either left or right-going waves.
As it will turn out, if  $\eta$ and $u$ are such that $\eta = K u$, 
then this pair of functions represents a solution
of \eqref{sys1}-\eqref{sys2} which is propagating to the right. 
%
% *************************************************************************************
Indeed, let us analyze the relation between $\eta$ and $u$ in the linearized Whitham system
\begin{align}
\label{linsys1}
\eta_t & = -  \mathcal{K}_0 u_x, \\
\label{linsys2}
u_t    & = - \eta_x.
\end{align}
%
Considering a solution of the system \eqref{linsys1}-\eqref{linsys2} in the form
\begin{equation}
\eta(x,t) = A e^{(i\xi x - i \omega t)}, \qquad u(x,t) = B e^{(i\xi x - i \omega t)}.
\end{equation}
%
\noindent 
gives rise to the matrix equation
\begin{equation}
\begin{pmatrix}
-\omega & \frac{\tanh{\xi}}{\xi} \xi \\
\xi  & -\omega
\end{pmatrix}
\begin{pmatrix}
A \\
B
\end{pmatrix} =
\begin{pmatrix}
0 \\
0
\end{pmatrix}.
\label{matrix-equation}
\end{equation}
%
If existence of a nontrivial solution of this system is to be guaranteed,
the determinant of the matrix has to be zero, so that we have
$\omega^2 - \frac{\tanh{\xi}}{\xi} \xi^2 = 0$. Defining the phase speed
as $c = \omega / \xi$, we obtain the dispersion relation
\begin{equation}
c = \pm \sqrt{{\textstyle \frac{\tanh{\xi}}{\xi}}}.
\end{equation}
%
%
The choice of $c > 0$ corresponds to right-going wave solutions of the system 
\eqref{linsys1}-\eqref{linsys2},
and the relation between $\eta$ and $u$ can be deduced from \eqref{linsys2}.
%
%
Accordingly, it is expedient to separate solutions
into a right-going part $r$ and a left-going part $s$
which are defined by
\begin{align*}
r = \ohf (\eta + K  u), \qquad
s = \ohf (\eta - K  u). 
\end{align*}
According to the transformation theory detailed in \cite{Craig2005}, 
if the unknowns $r$ and $s$ are used instead of $\eta$ and $u$,
the structure map changes to 
\begin{equation}
J_{r,s} = \left( \frac{\partial F}{\partial (\eta, u)} \right) J_{\eta, u}  
\left( \frac{\partial F}{\partial (\eta, u)} \right)^T
= 
\begin{pmatrix}
-\ohf \partial_x K & 0\\
0 & \ohf \partial_x K
\end{pmatrix}.
\end{equation}
%
\noindent 
We now use the same scaling for both dependent and independent variables as before.
Thus we have $r = \alpha\tr$ and $s = \alpha\ts$. 
%
%
%
% *************************************************************************************
%
The Hamiltonian is written in terms of $\tr$ and $\ts$ as
\begin{multline*}
\tH = \ohf \int_{\R} (\tr+\ts)^2 \, dx
\\
 + \ohf \int_{\R} \tK^{-1}(\tr-\ts) \left[ 1 - \Sfrac{1}{3} \mu^2 \tD^2 + \cdots \right] \tK^{-1}(\tr-\ts) \, dx
%\\
 + \frac{\alpha}{2} \int_{\R}  (\tr+\ts) \left(\tK^{-1}(\tr-\ts)\right)^2 \, dx
\\
 - \frac{\alpha}{2} \int_{\R}  \tK^{-1}(\tr-\ts) \left[ \mu \tD - \Sfrac{1}{3} \mu^3\tD^3 +  \cdots \right] (\tr+\ts)
   \left[ \mu \tD - \Sfrac{1}{3} \mu^3\tD^3 + \cdots \right] \tK^{-1}(\tr-\ts) \, dx. 
\end{multline*}
%
%
%
% *************************************************************************************
%
%
Following the transformation rules, the structure map transforms to
$J_{\tr,\ts} = 1/\ta^2 J_{r,s}$. 
In addition, the time scaling $t = \lambda \ttt$ is employed. 
Since the focus is on right-going solutions, the equation to be considered is
\begin{equation}
\lambda \tr_{\ttt} 
= -\frac{1}{2\ta^2}\lambda\partial_{\tx} \tK \left[ \frac{\delta \big( \ta^2 \tH \big)}{\delta \tr} \right].
\end{equation}
So far, this equation is exact. If we now assume that 
%the waves are predominantly right-going, and 
$s$ is of the order of
$\scrO(\mu^2 e^{\sca})$, then the equation for $\tr$ is
\begin{multline*}
\tr_{\ttt} = -\frac{1}{2}\partial_{\tx} \big[ 1 - \Sfrac{1}{6} \mu^2 \tD^2 + \cdots \big] 
           \bigg\{ 2 \tr + \frac{\alpha}{2} \left( \big[1 + \Sfrac{1}{6} \mu^2 \tD^2 + \cdots \big] 
           \tr \right)^2 
%\right.
\\
+ \alpha \big[ 1 + \Sfrac{1}{6} \mu^2 \tD^2 + \cdots \big] 
   \left( \tr \, \big[1 + \Sfrac{1}{6} \mu^2 \tD^2 + \cdots \big]  \tr  \right)  
- \frac{\alpha}{2} \left( \big[ \mu \tD - \Sfrac{1}{3} \mu^3 \tD^3 + \cdots \big] 
    \big[ 1 + \Sfrac{1}{6} \mu^2 \tD^2 + \cdots \big]  \tr  \right)^2
\\
%\left. 
- \alpha \big[ \mu \tD - \Sfrac{1}{3} \mu^3 \tD^3 + \cdots \big] \big[1 + \Sfrac{1}{6} \mu^2 \tD^2 + \cdots \big] 
 \left( \tr \big[ \mu \tD - \Sfrac{1}{3} \mu^3 \tD^3 + \cdots \big] [1 + \Sfrac{1}{6} \mu^2 \tD^2 
    + \cdots \big] \tr \right) \bigg\} + \scrO(\alpha \mu^2).
\end{multline*}
As in the case of the Whitham system, we use $\ta = \scrO(e^{- \kappa / \mu^\nu})$,
and disregard terms of order $\scrO(\mu^2 e^{\sca})$, but not of order $\scrO(e^{\sca})$.
This procedure yields the Whitham equation \eqref{whitham-equation} 
which is written in nondimensional variables as
\begin{equation*}
r_t = -K r_x - \frac{3}{2} r r_x.
\end{equation*}
As was the case for the system found in the previous section, it is also 
possible here to include terms of order $\scrO(\mu^2 e^{-\kappa/\mu^\nu})$, resulting
in the higher-order equation
\begin{equation*}
r_t = -K r_x - \frac{3}{2} r r_x - \frac{13}{12} r_x r_{xx} - \frac{5}{12} r r_{xxx}.
\end{equation*}
%
%
% *************************************************************************************
\subsection{Numerical results.}
% *************************************************************************************
%
In this section, the performance of the Whitham
equation as a model for surface water waves 
is compared to the KdV equation \eqref{kdv-equation}, the BBM equation \eqref{regularized-long-wave},
and the Pad\'e (2,2) equation \eqref{pade-2-2}.
For this purpose initial data are imposed, the Whitham, KdV, BBM, and Pad\'e equations are solved numerically,
and the solutions are compared to numerical solutions of the full Euler equations
with free-surface boundary conditions. We continue to work in normalized
variables, such as stated in the beginning of Section 2.

The numerical treatment of the three model equations is 
by a standard pseudo-spectral scheme,
such as explained in \cite{Ehrnstrom2013,Fornberg1978} for example. For the time stepping,
an efficient fourth-order implicit method developed in \cite{DeFrutos1992} is used. 
The numerical treatment of the free-surface problem for the Euler equations
is based on a conformal mapping of the fluid domain into a rectangle.
In the time-dependent case, this method has roots in the work of Ovsyannikov \cite{Ovsyannikov1974}, 
and was later used in \cite{Dyachenko1996} and \cite{Li2004}. 
The particular method used for the numerical experiments reported here 
is a pseudo-spectral scheme which is detailed in \cite{Mitsotakis2014}.
